Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2022-01-26 Thesis/02-chap2.Rmd
add paper 1 intro

% This is a LaTeX thesis template for Monash University.
% to be used with Rmarkdown
% This template was produced by Rob Hyndman
% Version: 6 September 2016

\documentclass{monashthesis}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Add any LaTeX packages and other preamble here if required
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author{Weihao Li}
\title{Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics}
\degrees{B.Comm. (Hons), Monash University}
\def\degreetitle{Doctor of Philosophy}
% Add subject and keywords below
\hypersetup{
     %pdfsubject={The Subject},
     %pdfkeywords={Some Keywords},
     pdfauthor={Weihao Li},
     pdftitle={Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics},
     pdfproducer={Bookdown with LaTeX}
}

\bibliography{thesisrefs}

\begin{document}

\pagenumbering{roman}

\titlepage

{\setstretch{1.2}\sf\tighttoc\doublespacing}

\hypertarget{copyright-notice}{%
\chapter*{Copyright notice}\label{copyright-notice}}
\addcontentsline{toc}{chapter}{Copyright notice}

\emph{(Choose one of the following notices.)}

\emph{(Notice 1)}

\textcopyright { } \authorname~(\number\the\year).

\emph{The second notice certifies the appropriate use of any third-party material in the thesis. Students choosing to deposit their thesis into the restricted access section of the repository are not required to complete Notice 2.}

\emph{(Notice 2)}

\textcopyright { } \authorname~(\number\the\year).

I certify that I have made all reasonable efforts to secure copyright permissions for third-party content included in this thesis and have not knowingly added copyright content to my work without the owner's permission.

\newpage

\hypertarget{abstract}{%
\chapter*{Abstract}\label{abstract}}
\addcontentsline{toc}{chapter}{Abstract}

The abstract should outline the main approach and findings of the thesis and must not be more than 500 words.

\newpage

\hypertarget{declaration}{%
\chapter*{Declaration}\label{declaration}}
\addcontentsline{toc}{chapter}{Declaration}

\emph{(Standard thesis)}

This thesis is an original work of my research and contains no material which has been accepted for the award of any other degree or diploma at any university or equivalent institution and that, to the best of my knowledge and belief, this thesis contains no material previously published or written by another person, except where due reference is made in the text of the thesis.

\emph{(Thesis including published works declaration)}

I hereby declare that this thesis contains no material which has been accepted for the award of any other degree or diploma at any university or equivalent institution and that, to the best of my knowledge and belief, this thesis contains no material previously published or written by another person, except where due reference is made in the text of the thesis.

This thesis includes (insert number) original papers published in peer reviewed journals and (insert number) submitted publications. The core theme of the thesis is (insert theme). The ideas, development and writing up of all the papers in the thesis were the principal responsibility of myself, the student, working within the (insert name of academic unit) under the supervision of (insert name of supervisor).

(The inclusion of co-authors reflects the fact that the work came from active collaboration between researchers and acknowledges input into team-based research.) Remove this paragraph for theses with sole-authored work

In the case of (insert chapter numbers) my contribution to the work involved the following:

I have / have not renumbered sections of submitted or published papers in order to generate a consistent presentation within the thesis.

\begin{table}
\centering\footnotesize\tabcolsep=0.12cm
\begin{tabular}{|p{1.2cm}|p{2cm}|p{1.8cm}|p{3.4cm}|p{3.5cm}|p{1.5cm}|}
\hline
\RaggedRight\textbf{Thesis chapter}  & 
\RaggedRight\textbf{Publication title}  & 
\RaggedRight\textbf{Status (published, in press, accepted or returned for revision)}  &  
\RaggedRight\textbf{Nature and \% of student contribution} & 
\RaggedRight\textbf{Co-author name(s), nature and \% of co-author’s contribution} &  \RaggedRight\textbf{Co-author(s), Monash student Y/N} \\ 
\hline
2 & xx & xx  & xx & xx & \multicolumn{1}{c|}{N}   \\
\hline
3 & xx & xx  & xx & xx  & \multicolumn{1}{c|}{N}   \\
\hline
4 & xx & xx  & xx & xx & \multicolumn{1}{c|}{N}    \\
\hline
5 & xx & xx  & xx & xx & \multicolumn{1}{c|}{N}   \\
\hline
\end{tabular}
\end{table}

\textbf{Student name:} \authorname

\textbf{Student signature:}

\textbf{Date:}

\hypertarget{acknowledgements}{%
\chapter*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{chapter}{Acknowledgements}

I would like to thank my pet goldfish for \dots

\hypertarget{preface}{%
\chapter*{Preface}\label{preface}}
\addcontentsline{toc}{chapter}{Preface}

The material in Chapter \ref{ch:intro} has been submitted to the journal \emph{Journal of Impossible Results} for possible publication.

The contribution in Chapter \ref{ch:litreview} of this thesis was presented in the International Symposium on Nonsense held in Dublin, Ireland, in July 2015.

\clearpage\pagenumbering{arabic}\setcounter{page}{0}

\hypertarget{ch:intro}{%
\chapter{Introduction}\label{ch:intro}}

\hypertarget{ch:paper1}{%
\chapter{Automatic Visual Statistical Inference, with Application to Linear Regression Diagnostics}\label{ch:paper1}}

\hypertarget{abstract-1}{%
\section{Abstract}\label{abstract-1}}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{model-diagnostics}{%
\subsection{Model Diagnostics}\label{model-diagnostics}}

@@

However, unlike confirmatory data analysis which is built upon rigorous statistical procedures, e.g., hypothesis testing, visual diagnostics relies on graphical perception - human's ability to interpret and decode the information embedded in the graph \autocite{cleveland_graphical_1984}, which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding a separation between gene groups in a two-dimensional projection from a linear discriminant analysis where there is no difference in the expression levels between the gene groups \autocite{roy_chowdhury_using_2015}.

\hypertarget{visual-inference}{%
\subsection{Visual Inference}\label{visual-inference}}

Visual inference was first introduced by \textcite{buja_statistical_2009} as an inferential framework to extend confirmatory statistics to visual discoveries. This framework redefines the test statistics, tests, null distribution, significance levels and \(p\)-value for visual discovery modelled on the confirmatory statistical testing. Figure \ref{fig:parallelism} outlines the parallelism between conventional tests and visual discovery.

\begin{figure}
\centering
\includegraphics[width=4.6875in,height=3.55208in]{figures/rsta2009012001.jpg}
\caption{Parallelism between multiple quantitative testing and visual discovery \autocite{buja_statistical_2009}. \label{fig:parallelism}}
\end{figure}

In visual inference, a visual discovery is defined as a rejection of a null hypothesis, and the same null hypothesis can be rejected by many different visual discoveries \autocite{buja_statistical_2009}. For model diagnostics, the null hypothesis would be the assumed model, while the visual discoveries would be any findings that are inconsistent with the hypothesis. The same assumed model, such as classical linear regression model, can be rejected by both nonlinearity and heteroskedasticity with the residual plot as shown in Figure \ref{fig:nonlinearityheter}.

\hypertarget{pre-specification-of-visual-discoverable-features}{%
\subsection{Pre-specification of Visual Discoverable Features}\label{pre-specification-of-visual-discoverable-features}}

As discussed in \textcite{buja_statistical_2009}, in the practice of model diagnostics, the range of possible visual discoveries is not pre-specified. In other words, people do not explicitly specify which one or more visual features they are looking for before the read of the diagnostic plot. This is concerning since conventional hypothesis testing always requires the pre-specification of the parameter space \(\Theta\) of the parameter of interest \(\theta \in \Theta\) to form a valid inferential procedure. To address this issue, a collection of test statistics \(T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)\) is defined, where \(\boldsymbol{\mathrm{y}}\) is the data and \(\boldsymbol{I}\) is a set of all possible visual features. \textcite{buja_statistical_2009} described each of the test statistics \(T^{(i)}(\boldsymbol{\mathrm{y}})\) as a measurement of the degree of presence of a visual feature. Alternatively, \textcite{majumder_validation_2013} avoids the use of visual features and defined the visual statistics \(T(.)\) as a mapping from a dataset to a data plot. Both definitions of visual test statistics are valid, but in the rest of the paper, the first definition will be used as it covered some details needed by this work.

The size of the collection \(T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)\) depends on the size of the set \(I\). If one can define \(I\) comprehensively, i.e, pre-specify all the visual discoverable features, the validity issue will be solved. Unfortunately, to our knowledge, there is no such a way to list all visual features. In linear regression diagnostics, possible visual features of a residual plot may be outliers, shapes and clusters. But these are very vague features which does not cover the full list of visual features.

Similarly, \textcite{wilkinson_graph-theoretic_2005} proposed the work called graph theoretic scagnostics, which adopted the idea of ``scagnostics'' - scatter plot diagnostics from . It includes 9 computable scagnostics measures defined on planar proximity graphs: ``Outlying'', ``Convex'', ``Skinny'', ``Stringy'', ``Straight'', ``Monotonic'', ``Skewed'', ``Clumpy'' and ``Striated'' which attempts to describe outliers, shape, density, trend and coherence of the data. This approach is inspiring but it does not contain the complete list of visual discoverable features. It is possible that such a list will never be complete as suggested in \textcite{buja_statistical_2009}.

Thinking out of the box, \textcite{buja_statistical_2009} argued that there is actually no need for pre-specification of visual discoverable features. In model diagnostics, when the null hypothesis is rejected, the reasons for rejecting the hypothesis will also be known. This is because observers not only can point out the fact that visual discoveries have been found, but also can describe the particular features they observed. Those features will correspond to the subset of the collection of visual test statistics \(T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)\) which resulted in rejection.

\begin{figure}
\centering
\includegraphics{thesis_files/figure-latex/nonlinearityheter-1.pdf}
\caption{\label{fig:nonlinearityheter}test}
\end{figure}

\hypertarget{lineup-protocol}{%
\subsection{Lineup Protocol}\label{lineup-protocol}}

\hypertarget{type-i-error}{%
\subsubsection{Type-I error}\label{type-i-error}}

\hypertarget{type-ii-error}{%
\subsubsection{Type-II error}\label{type-ii-error}}

\hypertarget{visual-inference-applied-to-linear-regression}{%
\subsection{Visual Inference Applied to Linear Regression}\label{visual-inference-applied-to-linear-regression}}

@@

\hypertarget{limitation-of-the-visual-inference}{%
\subsection{Limitation of the Visual Inference}\label{limitation-of-the-visual-inference}}

@@

\hypertarget{computer-vision-model}{%
\subsection{Computer Vision Model}\label{computer-vision-model}}

@@

\hypertarget{contribution}{%
\subsection{Contribution}\label{contribution}}

@@

\hypertarget{structure-of-this-paper}{%
\subsection{Structure of This Paper}\label{structure-of-this-paper}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  to check whether the assumptions has been violated
\item
  to check whether the line fit the data
\end{enumerate}

@@

\appendix

\hypertarget{additional-stuff}{%
\chapter{Additional stuff}\label{additional-stuff}}

You might put some computer output here, or maybe additional tables.

Note that line 5 must appear before your first appendix. But other appendices can just start like any other chapter.

\printbibliography[heading=bibintoc]

\end{document}

>>>>

---
chapter: 2
knit: "bookdown::render_book"
---

# Automatic Visual Statistical Inference, with Application to Linear Regression Diagnostics {#ch:paper1}

## Abstract

## Introduction

### Model Diagnostics

@@

However, unlike confirmatory data analysis which is built upon rigorous statistical procedures, e.g., hypothesis testing, visual diagnostics relies on graphical perception - human’s ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984], which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding a separation between gene groups in a two-dimensional projection from a linear discriminant analysis where there is no difference in the expression levels between the gene groups [@roy_chowdhury_using_2015].

### Visual Inference

Visual inference was first introduced by @buja_statistical_2009 as an inferential framework to extend confirmatory statistics to visual discoveries. This framework redefines the test statistics, tests, null distribution, significance levels and $p$-value for visual discovery modelled on the confirmatory statistical testing. Figure \ref{fig:parallelism} outlines the parallelism between conventional tests and visual discovery.

![(ref:parallelism) \label{fig:parallelism}](figures/rsta2009012001.jpg){width=450 height=341}

(ref:parallelism) Parallelism between multiple quantitative testing and visual discovery [@buja_statistical_2009].

In visual inference, a visual discovery is defined as a rejection of a null hypothesis, and the same null hypothesis can be rejected by many different visual discoveries [@buja_statistical_2009]. For model diagnostics, the null hypothesis would be the assumed model, while the visual discoveries would be any findings that are inconsistent with the hypothesis. The same assumed model, such as classical linear regression model, can be rejected by both nonlinearity and heteroskedasticity with the residual plot as shown in Figure \ref{fig:nonlinearityheter}.

### Pre-specification of Visual Discoverable Features

As discussed in @buja_statistical_2009, in the practice of model diagnostics, the range of possible visual discoveries is not pre-specified. In other words, people do not explicitly specify which one or more visual features they are looking for before the read of the diagnostic plot. This is concerning since conventional hypothesis testing always requires the pre-specification of the parameter space $\Theta$ of the parameter of interest $\theta \in \Theta$ to form a valid inferential procedure. To address this issue, a collection of test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$ is defined, where $\boldsymbol{\mathrm{y}}$ is the data and $\boldsymbol{I}$ is a set of all possible visual features. @buja_statistical_2009 described each of the test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})$ as a measurement of the degree of presence of a visual feature. Alternatively, @majumder_validation_2013 avoids the use of visual features and defined the visual statistics $T(.)$ as a mapping from a dataset to a data plot. Both definitions of visual test statistics are valid, but in the rest of the paper, the first definition will be used as it covered some details needed by this work.

The size of the collection $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$ depends on the size of the set $I$. If one can define $I$ comprehensively, i.e, pre-specify all the visual discoverable features, the validity issue will be solved. Unfortunately, to our knowledge, there is no such a way to list all visual features. In linear regression diagnostics, possible visual features of a residual plot may be outliers, shapes and clusters. But these are very vague features which does not cover the full list of visual features.

Similarly, @wilkinson_graph-theoretic_2005 proposed the work called graph theoretic scagnostics, which adopted the idea of "scagnostics" - scatter plot diagnostics from . It includes 9 computable scagnostics measures defined on planar proximity graphs: "Outlying", "Convex", "Skinny", "Stringy", "Straight", "Monotonic", "Skewed", "Clumpy" and "Striated" which attempts to describe outliers, shape, density, trend and coherence of the data. This approach is inspiring but it does not contain the complete list of visual discoverable features. It is possible that such a list will never be complete as suggested in @buja_statistical_2009.

Thinking out of the box, @buja_statistical_2009 argued that there is actually no need for pre-specification of visual discoverable features. In model diagnostics, when the null hypothesis is rejected, the reasons for rejecting the hypothesis will also be known. This is because observers not only can point out the fact that visual discoveries have been found, but also can describe the particular features they observed. Those features will correspond to the subset of the collection of visual test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$ which resulted in rejection.

```{r nonlinearityheter, fig.cap="test"}
library(ggplot2)
set.seed(10086)
cm <- visage::cubic_model(a = 0, b = 100, c = 1)

dat <- cm$gen(1000, fit_model = TRUE)
cm$plot(dat) +
  ggtitle("Residuals against fitted values plot of a classical linear regression model") +
  theme_light()
```

### Lineup Protocol

#### Type-I error

#### Type-II error

### Visual Inference Applied to Linear Regression

@@

### Limitation of the Visual Inference

@@

### Computer Vision Model

@@

### Contribution

@@

### Structure of This Paper

@@

1. to check whether the assumptions has been violated
2. to check whether the line fit the data

@@
